Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package

### 2019-12-05 doc/ihdbristol.Rmd
vignette tweaks

title: "Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package"
author: Chris Jackson <chris.jackson@mrc-bsu.cam.ac.uk>
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
    number_sections: true
vignette: >
 %\VignetteIndexEntry{Vignette Title}
 %\VignetteEngine{knitr::rmarkdown}
 \usepackage[utf8]{inputenc}
bibliography: disbayes.bib
---

@@

  * (incidence): $n_a^{(inc)}, r_a^{(inc)}$: given a population of size $n_a$^{(mort)}, $r_a^{(inc)}$ of these are observed to get the disease within the next year

  * (mortality): $n_a^{(mort)}, r_a^{(mort)}$: given a population of size $n_a^{(mort)}$ (with or without the disease), $r_a^{(mort)}$ of these are observed to die from the disease within the next year

  * (prevalence): $n_a^{(prev)}, r_a^{(prev)}$: from a sample of $n_a^{(prev)}$ individuals, $r_a^{(prev)}$ are known to have the disease at age $a$ (and $n_a^{(prev)} - r_a^{(prev)} are known to not have the disease.

@@

## Given estimates and denominators

@@

## Given point and interval estimates

@@

## Expressing additional uncertainty

@@

The `disbayes` package uses Stan to fit the Bayesian disease model given here to the data $\mathbf{y} = \{i_a, r_a, n_a\}$ supplied in the `ihdbristol` data frame.   The function `disbayes` does the work of converting the data to the required format for Stan, and initializing and a set of Markov Chain Monte Carlo simulations, and calling the Stan software to run the simulations.

@@

```{r, eval=TRUE}
library(ggplot2)
plot(dbres) +  ylab("Case fatality") + xlab("Age") + ylim(0,0.5) + xlim(40,100) + 
  geom_line(aes(y=dismod_cf), data=ihdbristol, col="purple")
```

@@

For chronic diseases in general:

@@

* Use data from larger areas as priors for quantities for smaller areas

* Borrow information between men/women?

@@

* Account for variations in incidence and prevalence with time as well as age.  DisMod II allows time trends to be modelled, but assumes that time trends are independent of age.  Relaxing this assumption would be feasible, but possibly computationally expensive. 
	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011.

@@

>>>>

---
title: "Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package"
author: Chris Jackson <chris.jackson@mrc-bsu.cam.ac.uk>
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Bayesian estimation of chronic disease epidemiology from incomplete data: the disbayes package}
 %\VignetteEngine{knitr::rmarkdown}
 \usepackage[utf8]{inputenc}
bibliography: disbayes.bib
---

```{r}
knitr::opts_chunk$set(fig.width=7)
```

@@

  * (incidence): given a population of size $n_a^{(mort)}$, $r_a^{(inc)}$ of these are observed to get the disease within the next year

  * (mortality): given a population of size $n_a^{(mort)}$ (with or without the disease), $r_a^{(mort)}$ of these are observed to die from the disease within the next year

  * (prevalence): from a sample of $n_a^{(prev)}$ individuals, $r_a^{(prev)}$ are known to have the disease at age $a$ (and $n_a^{(prev)} - r_a^{(prev)}$ are known to not have the disease.

@@

## ...Given estimates and denominators

@@

## ...Given point and interval estimates

@@

## ...Expressing additional uncertainty

@@

The `disbayes` package uses Stan to fit the Bayesian disease model given here to the data $\mathbf{y} = \{i_a, r_a, n_a\}$ supplied in the `ihdbristol` data frame.   The function `disbayes` does the work of converting the data to the required format for Stan, and calling the Stan software to run the simulations.

@@

```{r, eval=TRUE, warning=FALSE}
library(ggplot2)
plot(dbres) +  ylab("Case fatality") + xlab("Age") + ylim(0,0.5) + xlim(40,100) + 
  geom_line(aes(y=dismod_cf), data=ihdbristol, col="purple")
```

@@

@@

* Hierarchical modelling to account for variations between ordered socioeconomic categories / income groups

* Borrow information between men/women

@@

* Account for variations in incidence and prevalence with time as well as age.  DisMod II allows time trends to be modelled, but assumes that time trends are independent of age.  Relaxing this assumption would be feasible, but possibly computationally expensive. 
	* For IHD in the UK, data on time trends may come from the Health Survey for England, years 1993, 1994, 1988, 2003, 2006, 2011?

@@

## References
